Rule Synthesis for Rule-based Intrusion Detectors

### 2020-06-03 paper/paper.tex
minor changes
including argmin

@@

\begin{wrapfigure}[8]{r}{0.21\textwidth}
    \vspace{-4ex}  
    \centering
    \scalebox{0.9}{  
      \begin{tikzpicture}
        \begin{axis}[
            width = 0.3\textwidth,
            height = 0.2\textwidth,            
            bar width=3pt,
            scaled ticks=false,
            ymajorticks=false,
            ybar stacked,
            ymax=250,
            enlargelimits=0.09,
            clip=true,
            legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
            xtick=data,
            xticklabel style={rotate=45, font=\footnotesize},
            tick pos=left,
            nodes near coords,
            every node near coord/.append style={font=\footnotesize},
            nodes near coords align={vertical},
            symbolic x coords={Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec},]
            \addplot coordinates {(Jan, 114) (Feb, 155) (Mar, 117) (Apr, 159) (May, 110) (Jun, 231) (Jul, 122) (Aug, 167) (Sep, 107) (Oct, 225) (Nov, 148) (Dec, 127)};
        \end{axis}
      \end{tikzpicture}
    }
    \vspace{-5ex}
    \caption{\label{fig:distribution-rules-per-month}2019 new rules.}
\end{wrapfigure}
Open-source \nids{}, like Snort~\cite{snort} and
Suricata~\cite{suricata}, provide rulesets for many existing
attacks. Unfortunately, attackers actively look for new mechanisms to
exploit systems through the
network. Figure~\ref{fig:distribution-rules-per-month} shows the
number of rules added per month in the year of 2019 to the ``Emerging
Threats Open Ruleset''~\cite{emerging-threats-open}, a popular public
data set of Snort and \suri\ rules. A closer observation on the changelogs of
this ruleset~\cite{emerging-threats-changelogs} shows that new rules
are often created.  Rulesets need to be continuously revised as the
community finds and advertises vulnerabilities, which eventually can
be exploited by hackers.  Unfortunately, creating these rules is
time-consuming and error-prone~\cite{vollmer-etal-cics2011,alparslan-blog-suri}. The language for
describing rules is non-trivial (see
Section~\ref{sec:example-suricata-rules}) and testing correctness of a
rule requires data and careful inspection.

@@

\subsubsection{\label{sec:popularity-options}Popularity of options}

\tname{} use the popularity of the options used in trusted rules as a
heuristic function---rules containing popular options score higher
compared to those containing unpopular
options. Figure~\ref{fig:distribution-options} shows the histogram of
rule options for the ruleset we used~\cite{emerging-threats-open}. The
x-axis shows the options used in rules (the 8 most common) and the
y-axis shows the number of rules using the corresponding option. Note
that the options \CodeIn{content}, \CodeIn{flow}, and \CodeIn{prce}
are very popular. \tname\ uses the following formula to compute the
contribution of option popularity to the score of a rule $r$:
\indent
$\sum_{i=1}^{N}\frac{\mathit{option\_frequency[option_i]}}{\mathit{max(option\_frequency)}}/N$.

@@

\subsubsection{Number of content options} We observed
that \percContentOptions\ of the rules from the ruleset we analyzed
contain content options, but the distribution of those contents per
rule is not uniform. Figure~\ref{fig:distribution-contents} shows the
distribution of number of contents in a rule. Results indicate that
nearly 73\% of the rules from our ruleset contain one to three content
options. Hence, we found that that information could help
\tname\ discriminate useful rules. \tname{} computes the contribution
of the number of content options to the score of a rule $r$ with the
fraction
$\frac{\mathit{rules\_per\_number\_of\_contents[len(r)]}}{\mathit{max(rules\_per\_number\_of\_contents)}}$,
where $\mathit{len(r)}$ denotes the number of content options in $r$
and $\mathit{rules\_per\_number\_of\_contents}$ is the distribution of
number of contents. The expression in the numerator
%$\mathit{rules\_per\_number\_of\_contents[len(r)]}$
calculates the number of rules in the ruleset with the same number of
contents as $r$ whereas the denominator normalizes the value in the
0-1 range. To illustrate, the contribution of this function would be
$.64$~($=5841/9164$) and $.18$~($=1622/9164$) to the score of rules
with 3 and 5 options, respectively.

@@

\subsubsection{Popularity of content strings} 
This heuristic leverages content rarity to discriminate rules. It
weighs higher (respectively, lower) rules that use popular
(respectively, unpopular) strings in content options. The score of a
rule $r$ on this function is the arithmetic mean of the popularity
scores of each string used in contents in $r$. More precisely, content
rarity is obtained with the following formula:
\indent
$\sum_{i=1}^{N}\frac{\mathit{content\_frequency[content_i]}}{\mathit{max(content\_frequency)}}/N$

@@

\subsubsection{\label{sec:popularity-content-modifiers}Popularity of http content modifiers} As previously
mentioned, \percRulesWithContent\ of \suri\ rules involve content
options. Furthermore, nearly half of the rules that use content
options use content modifiers to qualify the search. Http modifiers
typically serve to specify parts of the message payloads that the
input string should be matched against. For example, the modifier
\CodeIn{http\_header} in the option \CodeIn{content: ``keep-alive'';
  http\_header} indicates that the string ``keep-alive'' should be
present in the header of the http message. (A keep-alive connection
attribute indicates that a single tcp connection should remain open
for multiple http requests/responses.) In this case, the content
modifier refers to an http field that appears in the payload of a tcp
message, which is the target of the option
\CodeIn{content}. Figure~\ref{fig:http-header-example} illustrates
possible content options used to match against certain strings that
appear in an http message. A check mark (respectively, cross)
indicates a match (respectively, no match).
\tname\ builds on the observation that the strings commonly accessed
at positions associated with content modifiers (\eg{}, host,
connection) are more common. For example, historical data indicates
that content options that use the modifier \CodeIn{connection} are
much less likely than content options that use the modifier
\CodeIn{uri}. Figure~\ref{fig:distribution-content_modifiers} shows
the histogram of content modifiers for http. (We removed the prefix
\CodeIn{http} on the modifier for space.) Although content modifiers
are applicable to rules of any protocol supported by Suricata,
currently, \tname\ only supports this features for tcp and http; these
rules comprise $\sim$80\% of the rules of our data set. We compute the
heuristic function for content modifiers with the following formula:

@@

\subsubsection{Weight Optimization}
\label{sec:weight-optimization}
A key strategy for obtaining the ideal ranking lie in finding optimal weights
of these heuristic functions. 
%We ran \tname{} using different combinations of weights on a sample of
%attacks and chose the combination that offered the best overall
%performance, which we computed with the sum of the rankings of the
%golden rule on each attack. The weight combination that led to the
%smaller sum of rankings was considered the best choice. We considered
%\emph{all} combinations of the following weights 0, 0.25, 0.5, 0.75,
%and 1. Given that there are four different functions and five possible
%weight values, we needed to run \tname{} on the sample of attacks for
%625 (=$5^4$) times. In the end of the process, the weight combination
%$\langle{}1,0.25,0.25,1\rangle$ was the best choice we found. Although
%results showed that this strategy results in good performance, we
%remain to evaluate other optimization strategies to find better
%weights.
%
In  order to compute the optimal weights, we used the Differential Evolution 
algorithm from SciPy\cite{diff-evo-scipy} to find the optimal weights for our set of attacks.
The optmization consists in minimizing the function that returns the sum of 
every attack's golden rule ranking in our set given a combination of weights: 
\indent
$\sum_{i=1}^{N}\mathit{R_i}$ ($\mathit{R_i}$, where $\mathit{R}$ is the 
ranking of a golden rule and $\mathit{N}$ is the number of attacks).

The algorithm also takes as input the range of the weights (between 0 and 1) 
and the max number of generations, which we set to 10 (confirmation whether 
this number suffices remains to future work). The weights are 
as follows $\langle{}0.280, 0.057, 0.043, 0.490\rangle$.

>>>>

%
% Math stuff
%
\newcommand{\argmin}{\mathop{\mathrm{{arg\,min}}}}

@@

\begin{wrapfigure}[7]{r}{0.21\textwidth}
    \vspace{-4ex}  
    \centering
    \scalebox{0.9}{  
      \begin{tikzpicture}
        \begin{axis}[
            width = 0.3\textwidth,
            height = 0.2\textwidth,            
            bar width=3pt,
            scaled ticks=false,
            ymajorticks=false,
            ybar stacked,
            ymax=250,
            enlargelimits=0.09,
            clip=true,
            legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
            xtick=data,
            xticklabel style={rotate=45, font=\footnotesize},
            tick pos=left,
            nodes near coords,
            every node near coord/.append style={font=\footnotesize},
            nodes near coords align={vertical},
            symbolic x coords={Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec},]
            \addplot coordinates {(Jan, 114) (Feb, 155) (Mar, 117) (Apr, 159) (May, 110) (Jun, 231) (Jul, 122) (Aug, 167) (Sep, 107) (Oct, 225) (Nov, 148) (Dec, 127)};
        \end{axis}
      \end{tikzpicture}
    }
    \vspace{-5ex}
    \caption{\label{fig:distribution-rules-per-month}2019 new rules.}
\end{wrapfigure}
Open-source \nids{}, like Snort~\cite{snort} and
Suricata~\cite{suricata}, provide rulesets for many existing
attacks. Unfortunately, attackers actively look for new mechanisms to
exploit systems through the
network. Figure~\ref{fig:distribution-rules-per-month} shows the
number of rules added per month in the year of 2019 to the ``Emerging
Threats Open Ruleset''~\cite{emerging-threats-open}, a popular public
data set of Snort and \suri\ rules. A closer observation on the changelogs of
this ruleset~\cite{emerging-threats-changelogs} shows that new rules
are often created.  Rulesets need to be continuously revised as the
community finds and advertises vulnerabilities, which eventually can
be exploited by hackers.  Unfortunately, creating these rules is
time-consuming and error-prone~\cite{vollmer-etal-cics2011,alparslan-blog-suri}. The language for
describing rules is non-trivial (see
Section~\ref{sec:example-suricata-rules}) and testing correctness of a
rule requires data and careful inspection.

@@

\subsubsection{\label{sec:popularity-options}Popularity of options ($popt$)}

\tname{} use the popularity of the options used in trusted rules as a
heuristic function---rules containing popular options score higher
compared to those containing unpopular
options. Figure~\ref{fig:distribution-options} shows the histogram of
rule options for the ruleset we used~\cite{emerging-threats-open}. The
x-axis shows the options used in rules (the 8 most common) and the
y-axis shows the number of rules using the corresponding option. Note
that the options \CodeIn{content}, \CodeIn{flow}, and \CodeIn{prce}
are very popular. \tname\ uses the following formula to compute the
contribution of option popularity to the score of a rule $r$:
\indent
$popt = \sum_{i=1}^{N}\frac{\mathit{option\_frequency[option_i]}}{\mathit{max(option\_frequency)}}/N$.

@@

\subsubsection{Number of content options ($ncont$)} 
We observed
that \percContentOptions\ of the rules from the ruleset we analyzed
contain content options, but the distribution of those contents per
rule is not uniform. Figure~\ref{fig:distribution-contents} shows the
distribution of number of contents in a rule. Results indicate that
nearly 73\% of the rules from our ruleset contain one to three content
options. Hence, we found that that information could help
\tname\ discriminate useful rules. \tname{} computes the contribution
of the number of content options to the score of a rule $r$ with the
fraction
$\frac{\mathit{rules\_per\_number\_of\_contents[len(r)]}}{\mathit{max(rules\_per\_number\_of\_contents)}}$,
where $\mathit{len(r)}$ denotes the number of content options in $r$
and $\mathit{rules\_per\_number\_of\_contents}$ is the distribution of
number of contents. The expression in the numerator
%$\mathit{rules\_per\_number\_of\_contents[len(r)]}$
calculates the number of rules in the ruleset with the same number of
contents as $r$ whereas the denominator normalizes the value in the
0-1 range. To illustrate, the contribution of this function would be
$.64$~($=5841/9164$) and $.18$~($=1622/9164$) to the score of rules
with 3 and 5 options, respectively.

@@

\subsubsection{Popularity of content strings ($pcont$)} 
This heuristic leverages content rarity to discriminate rules. It
weighs higher (respectively, lower) rules that use popular
(respectively, unpopular) strings in content options. The score of a
rule $r$ on this function is the arithmetic mean of the popularity
scores of each string used in contents in $r$. More precisely, content
rarity is obtained with the following formula:
\indent
$\sum_{i=1}^{N}\frac{\mathit{content\_frequency[content_i]}}{\mathit{max(content\_frequency)}}/N$

@@

\subsubsection{\label{sec:popularity-content-modifiers}Popularity of http content modifiers ($phttp$)} 
As previously mentioned, \percRulesWithContent\ of \suri\ rules involve content
options. Furthermore, nearly half of the rules that use content
options use content modifiers to qualify the search. Http modifiers
typically serve to specify parts of the message payloads that the
input string should be matched against. For example, the modifier
\CodeIn{http\_header} in the option \CodeIn{content: ``keep-alive'';
  http\_header} indicates that the string ``keep-alive'' should be
present in the header of the http message. (A keep-alive connection
attribute indicates that a single tcp connection should remain open
for multiple http requests/responses.) In this case, the content
modifier refers to an http field that appears in the payload of a tcp
message, which is the target of the option
\CodeIn{content}. Figure~\ref{fig:http-header-example} illustrates
possible content options used to match against certain strings that
appear in an http message. A check mark (respectively, cross)
indicates a match (respectively, no match).
\tname\ builds on the observation that the strings commonly accessed
at positions associated with content modifiers (\eg{}, host,
connection) are more common. For example, historical data indicates
that content options that use the modifier \CodeIn{connection} are
much less likely than content options that use the modifier
\CodeIn{uri}. Figure~\ref{fig:distribution-content_modifiers} shows
the histogram of content modifiers for http. (We removed the prefix
\CodeIn{http} on the modifier for space.) Although content modifiers
are applicable to rules of any protocol supported by Suricata,
currently, \tname\ only supports this features for tcp and http; these
rules comprise $\sim$80\% of the rules of our data set. We compute the
heuristic function for content modifiers with the following formula:

@@

\subsubsection{Weight Optimization}
\label{sec:weight-optimization}
A key strategy for obtaining the ideal ranking lie in finding optimal weights
for the heuristic functions. The ranking of a golden rule, for a set of $N$ 
attacks, is given by $R(W) = \sum_{i=1}^{N}\mathit{R_i}$, where 
$R_i = (w_1 \cdot popt) \cdot (w_2 \cdot ncont) \cdot (w_3 \cdot pcont) \cdot (w_4 \cdot phttp)$
and $W = \langle{}w_1, w_2, w_3, w_4\rangle{}$ are the weights considered for each heuristic.
%
In  order to compute the optimal weights, we used the Differential Evolution 
algorithm from SciPy\cite{diff-evo-scipy}.
The optmization consists in minimizing the function of every attack's golden rule 
ranking in our set given a combination of weights. Formally, $W  = \displaystyle \argmin_W R$.
%
The algorithm also takes as input the range of the weights (between 0 and 1) 
and the max number of generations, which we set to 10 (confirmation whether 
this number suffices remains to future work). The weights are 
as follows $\langle{}0.280, 0.057, 0.043, 0.490\rangle{}$.
